Dynamic Witnesses for Static Type Errors

### 2015-11-20 paper/pldi16-submission/interactive.tex
kill subterm
more WRW comments

\subsection{Tracing Semantics}
\label{sec:inter-semant}
%
\begin{figure*}[t]
%% \relDescription{Computing Sub-Terms}
%% \begin{gather*}
%% \begin{array}{lcl}
%% \subtermssym                 & \dcolon & e \to \tr \\
%% \subterms{\eapp{e_1}{e_2}}   & \defeq & \subterm{\eapp{e_1}{e_2}}{e_1}; \subterm{\eapp{e_1}{e_2}}{e_2} \\
%% \subterms{\eplus{e_1}{e_2}}   & \defeq & \subterm{\eplus{e_1}{e_2}}{e_1}; \subterm{\eplus{e_1}{e_2}}{e_2} \\
%% \subterms{\eif{e_1}{e_2}{e_3}}   & \defeq & \subterm{\eif{e_1}{e_2}{e_3}}{e_1}; \\
                                %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_2}; \\
                                %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_3} \\
%% \subterms{\elet{x}{e_1}{e_2}}   & \defeq & \subterm{\elet{x}{e_1}{e_2}}{e_1}; \\
                                %% &        & \subterm{\elet{x}{e_1}{e_2}}{e_2} \\
%% \subterms{\efun{x}{e}}       & \defeq & \subterm{\efun{x}{e}}{e} \\
%% \subterms{e}                 & \defeq & \bullet
%% \end{array}
%% \end{gather*}
\judgementHead{Traced Evaluation}{\stepg{e}{\su}{\tr}{e}{\su}{\tr}}
\begin{gather*}
\inference[\recontext]
  {\stepg{e}{\su}{\tr}{e_1}{\su_1}{\tr_1}}
  {\stepg{C[e]}{\su}{\tr}{C[e_1]}{\su_1}{\singlestep{C[e]}{C[e_1]}; \subterms{C[e_1]}; \tr_1}}
\\ \\
\inference[\reappgood]
  {\pair{\efun{x}{e}}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}
         {e\sub{x}{v_2}}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \subterms{e\sub{x}{v_2}}; \tr}}
\\ \\
\inference[\reappbad]
  {\pair{\stuck}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}{\stuck}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{\stuck}; \tr}}
\end{gather*}
\caption{A selection of the operational semantics from
  Figure~\ref{fig:operational}, extended to collect a full reduction
  % WRW: \subterms is Used here in the figure before being Defined (indeed,
  % it is never defined anywhere that I can see). A simple bit in text to
  % explain what you mean would help. Why is subterms necessary? What does
  % it collect up for you? 
  graph.}
\label{fig:interactive}
\end{figure*}

\paragraph{Reduction Graphs}
%
A \emph{steps-to} edge is a pair of expressions \singlestep{e_1}{e_2}, which
% WRW: This was "e_1 steps, in a single step" which was circular.
intuitively indicates that $e_1$ reduces, in a single step, to $e_2$.
%
%A \emph{sub-term} edge is a pair of expressions \subterm{e_1}{e_2}, which
%intuitively indicates that $e_1$ contains $e_2$ as a sub-expression.
%
A \emph{reduction graph} is a set of steps-to edges:
$$\tr ::= \bullet \spmid \singlestep{e}{e}; \tr$$  %% \spmid \subterm{e}{e}; \tr$$

@@

\begin{figure}[t]
\centering
\includegraphics[width=\linewidth]{fac-steps.png}
\caption{A sequence of interactions with the trace of
  \texttt{fac 1}. The stuck term is red, in each node the redex is
  highlighted. Thick arrows denote a multi-step transition, thin arrows
  denote a single-step transition. We start in step 1. In step 2 we jump
  forward from the witness to the next function call. In step 3 we step
  into the recursive \texttt{fac 0} call, which spawns a new ``thread''
  of execution. In step 4 we take a single step forward from
  \texttt{fac 0}.}
\label{fig:nanomaly-factorial}
\end{figure}

>>>>

\subsection{Tracing Semantics}
\label{sec:inter-semant}
%
\begin{figure*}[t]
%% \relDescription{Computing Sub-Terms}
%% \begin{gather*}
%% \begin{array}{lcl}
%% \subtermssym                 & \dcolon & e \to \tr \\
%% \subterms{\eapp{e_1}{e_2}}   & \defeq & \subterm{\eapp{e_1}{e_2}}{e_1}; \subterm{\eapp{e_1}{e_2}}{e_2} \\
%% \subterms{\eplus{e_1}{e_2}}   & \defeq & \subterm{\eplus{e_1}{e_2}}{e_1}; \subterm{\eplus{e_1}{e_2}}{e_2} \\
%% \subterms{\eif{e_1}{e_2}{e_3}}   & \defeq & \subterm{\eif{e_1}{e_2}{e_3}}{e_1}; \\
                                %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_2}; \\
                                %% &        & \subterm{\eif{e_1}{e_2}{e_3}}{e_3} \\
%% \subterms{\elet{x}{e_1}{e_2}}   & \defeq & \subterm{\elet{x}{e_1}{e_2}}{e_1}; \\
                                %% &        & \subterm{\elet{x}{e_1}{e_2}}{e_2} \\
%% \subterms{\efun{x}{e}}       & \defeq & \subterm{\efun{x}{e}}{e} \\
%% \subterms{e}                 & \defeq & \bullet
%% \end{array}
%% \end{gather*}
\judgementHead{Traced Evaluation}{\stepg{e}{\su}{\tr}{e}{\su}{\tr}}
\begin{gather*}
\inference[\recontext]
  {\stepg{e}{\su}{\tr}{e_1}{\su_1}{\tr_1}}
  {\stepg{C[e]}{\su}{\tr}{C[e_1]}{\su_1}{\singlestep{C[e]}{C[e_1]}; \tr_1}}
\\ \\
\inference[\reappgood]
  {\pair{\efun{x}{e}}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}
         {e\sub{x}{v_2}}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{e\sub{x}{v_2}}; \tr}}
\\ \\
\inference[\reappbad]
  {\pair{\stuck}{\su_2} = \force{v_1}{\tfun{\thole{}}{\thole{}}}{\su_1}}
  {\stepg{\eapp{v_1}{v_2}}{\su_1}{\tr}{\stuck}{\su_1;\su_2}{\singlestep{\eapp{v_1}{v_2}}{\stuck}; \tr}}
\end{gather*}
\caption{A selection of the operational semantics from
  Figure~\ref{fig:operational}, extended to collect a full reduction
  graph.}
\label{fig:interactive}
\end{figure*}

\paragraph{Reduction Graphs}
%
A \emph{steps-to} edge is a pair of expressions \singlestep{e_1}{e_2}, which
intuitively indicates that $e_1$ reduces, in a single step, to $e_2$.
%
%A \emph{sub-term} edge is a pair of expressions \subterm{e_1}{e_2}, which
%intuitively indicates that $e_1$ contains $e_2$ as a sub-expression.
%
A \emph{reduction graph} is a set of steps-to edges:
$$\tr ::= \bullet \spmid \singlestep{e}{e}; \tr$$  %% \spmid \subterm{e}{e}; \tr$$

@@

\begin{figure}[t]
\centering
\includegraphics[width=\linewidth]{fac-steps-new.png}
\caption{A sequence of interactions with the trace of
  \texttt{fac 1}. The stuck term is red, in each node the redex is
  highlighted. Thick arrows denote a multi-step transition, thin arrows
  denote a single-step transition. We start in step 1. In step 2 we jump
  forward from the witness to the next function call. In step 3 we step
  into the recursive \texttt{fac 0} call, which spawns a new ``thread''
  of execution. In step 4 we take a single step forward from
  \texttt{fac 0}.}
\label{fig:nanomaly-factorial}
\end{figure}

### 2015-11-20 paper/pldi16-submission/semantics.tex
more WRW comments

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t} by
yielding a pair of either the same expression and an empty value
substitution, or yields the stuck term if no such refinement is
possible. In the case where \texttt{e} is a hole, it first checks in the
given $\vsu$ to see if the hole has already been instantiated and, if
so, returns the existing instantiation:
% WRW thinks we should explain why \force(\ehole,t,sigma) 
% is just sigma(i) and not \force(sigma(i),t, sigma) -- at this
% point the reader may not understand, especially since just two paragraphs
% above you made a big deal about sigma mapping holes to things that
% contain other holes.
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \pair{\vsu(i)}{\emptysu}        & \mbox{if}\ i \in \vsu \\
  \pair{v}{\ehole{i} \mapsto v}   & v = \gen{t}\\
\end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$

@@

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
% \ES{self-todo: take a close look at the $\trace$'s vs k's}
\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$ and
  $\trace' = \pair{\eapp{f}{\ehole{}}}{\emptysu},...,\pair{e'}{\vsu'}$ 
  such that $\step{e'}{\vsu'}{e}{\vsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
  in at most $k$ steps:
%
  %\begin{description}
  % \item
  [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  %\item
  [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k$.
    %
    % WRW suggests one more sentence here: do the thinking for the reader.
    % Why does t being incompatible with s_\tau mean that narrow
    % cannot succeed? I think officially you need to refer to Preservation
    % here in your induction, to argue that the type being narrowed *to*
    % is the same as the types you introduced earlier, and *thus* the
    % narrow must fail (because they're incompatible). Otherwise readers
    % might imagine a weird case where t and s_\tau are incompatible, but
    % the final narrow is to some other type t'' that is compatible. 
    %
    % This is a primary result, so more detail may not hurt.  
    A case analysis of the rules shows that \forcesym cannot have
    succeeded.
  %\end{description}
\end{proof}

@@

\paragraph{Generating Target Arguments}
%
We address the issue of currying by defining a procedure @genArgs@ such that \\
\hbox{@genArgs [(f1,e1),...,(fn,en)]@} yields a list of holes @[v1,...,vk]@
such that \hbox{@fn v1 ... vk@} \emph{does not} evaluate to a lambda.
%
This is achieved via a simple @loop@ that keeps adding holes until
% WRW: Where / to what does the loop add holes? Be specific for the reader. 
@eval@uating the closed term yields a non-lambda value.
%
The helper functions @witness@ and @close@ are used to respectively
apply the parameters to the target and close the result under the top-level
let-binders, prior to invoking \hbox{@eval@.}
%
That is, @mkApps@ creates a nested sequence of applications in
the usual left-associative style, and @mkLets@ takes a list of
binders and a body expression, and creates a sequence of nested
let-binders that close the body expression.

>>>>

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t} by
yielding a pair of either the same expression and an empty value
substitution, or yields the stuck term if no such refinement is
possible. In the case where \texttt{e} is a hole, it first checks in the
given $\vsu$ to see if the hole has already been instantiated and, if
so, returns the existing instantiation:
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex}
\begin{cases}
  \pair{\vsu(i)}{\emptysu}        & \mbox{if}\ i \in \vsu \\
  \pair{v}{\ehole{i} \mapsto v}   & v = \gen{t}\\
\end{cases} \\
 % \begin{array}{l}
    % \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
         % \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  % \end{array} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$
%
While a hole may map to a value that \emph{contains} another hole, \eg a
lambda, it may not map \emph{directly} to another hole, so in the second
case above we do not need to \forcesym\ the result of the substitution.

@@

\paragraph{Incompatible Values Are Wrong}
%
\emph{Any} value that is \emph{incompatible} with
the partial type upto trace $\trace$ will cause $f$ to get stuck
in \emph{at most} $k$ steps where $k$ is the length of $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu} and
     \hastype{v}{t} and
     \tincompat{t}{\ptype{\trace}{f}},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$ and
  $\trace' = \pair{\eapp{f}{\ehole{}}}{\emptysu},...,\pair{e'}{\vsu'}$ 
  such that $\step{e'}{\vsu'}{e}{\vsu}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
  in at most $k$ steps:
%
  %\begin{description}
  % \item
  [Case \tincompat{t}{s_{\trace'}}:]
    The inductive hypothesis applies.
  %\item
  [Case $\tcompat{t}{s_{\trace'}}$ but $\tincompat{t}{s_{\trace}}$:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k$.
    %
    Lemma~\ref{lem:refine-partial} tells us
    $\tcompat{s_{\trace'}}{s_{\trace}}$, which means we must have
    specifically narrowed $s_{\trace'}$ to a type incompatible with $t$.
    %
    A case analysis of the rules shows that \forcesym cannot have
    succeeded.
  %\end{description}
\end{proof}

@@

\paragraph{Generating Target Arguments}
%
We address the issue of currying by defining a procedure @genArgs@ such that \\
\hbox{@genArgs [(f1,e1),...,(fn,en)]@} yields a list of holes @[v1,...,vk]@
such that \hbox{@fn v1 ... vk@} \emph{does not} evaluate to a lambda.
%
This is achieved via a simple @loop@ that keeps adding holes to the
target application until @eval@uating the closed term yields a
non-lambda value.
%
The helper functions @witness@ and @close@ are used to respectively
apply the parameters to the target and close the result under the top-level
let-binders, prior to invoking \hbox{@eval@.}
%
That is, @mkApps@ creates a nested sequence of applications in
the usual left-associative style, and @mkLets@ takes a list of
binders and a body expression, and creates a sequence of nested
let-binders that close the body expression.
